---
title: "Assignment 7: Iteration (for loops) and functions"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Instructions: please read through this before you begin**

* This homework is due by **10pm on Monday 11/09/20.**

* Please **reproduce this markdown template.** Pay attention to all the formatting in this file, including bullet points, bolded characters, inserted code chuncks, headings, text colors, blank lines, and etc.

* When a verbal response is needed, answer by editing the part in the R markdown template where it says "Write your response here".

* Have all your code embedded within the R markdown file, and show both your code and plots in the knitted markdown file.

* Use R Markdown functionalities to **hide messages and warnings when needed**. (Suggestion:messages and warnings can often be informative and important, so please examine them carefully and only turn them off when you finish the exercise).

* Please name your R markdown file `assignment_7.Rmd` and the knitted markdown file `assignment_7.md`. Please upload both files using your personal GitHub repository for this class.


**Acknowledgments:** Exercise 1 has been adapted (with permission) from the **[Data Carpentry for Biologists semester course](https://datacarpentry.org/semester-biology/exercises/Functions-use-and-modify-Python/)** and exercises 2 and 3 are adapted (with permission) from lain Carmichael's course **[STOR 390:Introduction Data Science](https://idc9.github.io/stor390/#course_material).**

To start, first load all the required packages with the following code. Install them if they are not installed yet.


```{r, message = FALSE}

library(tidyverse)

```


### **Exercise 1: Body mass estimation using vectorization vs. for loop**

There are two major types of approaches to perform multiple operations in R:**[vectorization](https://swcarpentry.github.io/r-novice-gapminder/09-vectorization/)** and for loops. As a simple to calculate the sum of two vectors, `x` and `y`, the syntax for vectorization is simply `z <- x + y`. With this, the computer will be able to perform the same operation to each element of x and y vector **simultaneously.**

The for loop approach to achieve the same task (create a new vector z that is the sum of vectors x and y), on the other hand, takes the following form:

```{r, eval=FALSE}

z[i] <- NULL

for (i in 1:length(x)){
  
  z[i] <- x[i] +y[i]
  
  }

```

In this case, the computer loops through each element of x and y and performs the operation **sequentially**, resulting in a significantly longer runtime. Let's now try to quantify this difference in runtime in this exercise.




**1.1**

The length of an organism is typically strongly correlated with its body mass. This is useful because it allows us to estimate the mass of an organism even if we only know its length. This relationship generally takes the form: `mass = a * length ^ b`, where the parameters `a` and `b` vary among groups. This allometric approach is regularly used to estimate the mass of dinosaurs since we cannot weigh something that is only preserved as bones.

*Spinosaurus* is a predator that is bigger, and therefore, by definition, cooler, than that stupid *Tyrannosaurus* that everyone likes so much. It has an estimated `a` of `0.73` and `b` of `3.63.` What is the estimated mass of a *Spinosaurus* that is `16` m long based `16` m long based on its reassembled skeleton?


```{r}
##[1] 17150.56
```


 **1.2**
 
 The following vectors contain the `length`s of 40 dinosaurs and their respective `a` and `b` values. Estimate their `mass` first using a vectorization approach and then using a for and then using a for loop approach.
 
 
```{r,eval=FALSE}
dinosaur_lengths <- c(17.8013631070471, 20.3764452071665, 14.0743486294308, 25.65782386974, 26.0952008049675, 20.3111541103134, 17.5663244372533, 11.2563431277577, 20.081903202614, 18.6071626441984, 18.0991894513166, 23.0659685685892, 20.5798853467837, 25.6179254233558, 24.3714331573996, 26.2847248252537, 25.4753783544473, 20.4642089867304, 16.0738256364701, 20.3494171706583, 19.854399305869, 17.7889814608919, 14.8016421998303, 19.6840911485379, 19.4685885050906, 24.4807784966691, 13.3359960054899, 21.5065994598917, 18.4640304608411, 19.5861532398676, 27.084751999756, 18.9609366301798, 22.4829168046521, 11.7325716149514, 18.3758846100456, 15.537504851634, 13.4848751773738, 7.68561192214935, 25.5963348603783, 16.588285389794)

a_values <- c(0.759, 0.751, 0.74, 0.746, 0.759, 0.751, 0.749, 0.751, 0.738, 0.768, 0.736, 0.749, 0.746, 0.744, 0.749, 0.751, 0.744, 0.754, 0.774, 0.751, 0.763, 0.749, 0.741, 0.754, 0.746, 0.755, 0.764, 0.758, 0.76, 0.748, 0.745, 0.756, 0.739, 0.733, 0.757, 0.747, 0.741, 0.752, 0.752, 0.748)

b_values <- c(3.627, 3.633, 3.626, 3.633, 3.627, 3.629, 3.632, 3.628, 3.633, 3.627, 3.621, 3.63, 3.631, 3.632, 3.628, 3.626, 3.639, 3.626, 3.635, 3.629, 3.642, 3.632, 3.633, 3.629, 3.62, 3.619, 3.638, 3.627, 3.621, 3.628, 3.628, 3.635, 3.624, 3.621, 3.621, 3.632, 3.627, 3.624, 3.634, 3.621)

a_values*dinosaur_lengths^b_values

```

* with vectorization:

```{r}

##[1]  26039.686  42825.603  10800.224  98273.049 104257.481  41822.386  24840.644   4899.022  39915.948
#[10]  30937.922  26354.908  66384.865  43837.944  97141.451  80553.856 105556.405  97374.660  42760.136
##[19]  18749.274  42109.012  40674.182  26003.425  13229.824  37472.789  34684.033  80187.272   9460.977
##[28]  51630.571  29253.772  36399.306 117511.962  33384.288  58581.226   5462.316  28637.745  15864.172
##[37]   9284.810   1218.755  98522.609  19534.524

```


Hint: if you are unsure about how vectorized operations work, check out the results that the following lines return

```{r, eval = FALSE}

c(1, 2, 3) + c(4, 5, 6)
c(1, 2, 3) * c(4, 5, 6)

```


```{r, eval = FALSE}
mass <-  NULL

for (i in 1:length(dinosaur_lengths)){
  mass[i] <- a_values[i]*dinosaur_lengths[i]^b_values[i]
}
mass
```

* with a for loop

```{r}

## [1]  26039.686  42825.603  10800.224  98273.049 104257.481  41822.386  24840.644   4899.022  39915.948
##[10]  30937.922  26354.908  66384.865  43837.944  97141.451  80553.856 105556.405  97374.660  42760.136
##[19]  18749.274  42109.012  40674.182  26003.425  13229.824  37472.789  34684.033  80187.272   9460.977
##[28]  51630.571  29253.772  36399.306 117511.962  33384.288  58581.226   5462.316  28637.745  15864.172
##[37]   9284.810   1218.755  98522.609  19534.524




```



**1.3**

The function `system.time()` can be used to record the runtime of an operation. For example, `system.time(rnorm(100000))` can give you the time to draw 100000 samples from a normal distribution. Here, use `system.time()` to record the runtime of the vectorization approach and the for loop approach and the for loop approach in the previous question.


* vectorization: 

```{r, eval=FALSE, echo=FALSE}
system.time(a_values*dinosaur_lengths^b_values)
```

```{r}
 ##  user  system elapsed 
 ##  0       0       0 
```



* for loop:

```{r, eval = FALSE, echo=FALSE}

mass <-  NULL

system.time(for (i in 1:length(dinosaur_lengths)){
  mass[i] <- a_values[i]*dinosaur_lengths[i]^b_values[i]
})
mass

```


```{r}
## user  system elapsed 
## 0.004   0.000   0.004 
```


Although the for loop in this excercise can be run quickly, it is noticeably slower than the vectorization approach. With more complicated operations, vectorization can often shorten the runtime of a for loop from days to minutes. 

### **Exercise 2: Data inputting and wrangling in batch**

Another instance where loops are useful is data input/output in batch. We've learned how to make plots and output them in batch in class, so in this exercise, you will use for loops to automate the inputting and wrangling process of a group of datasets with similar names and formats in the https://github.com/nt246/NTRES6940-data-science/tree/master/datasets/buoydata.


**2.1 Given the following code chunk for reading buoy data files from buoy 44013 for each year, describe the following:**

* What parts of your code are consistent across every line/code chunk?

* What parts are different?


```{r, message = FALSE}

buoy_1987 <- read_csv('https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1987.csv', na = c("99", "999", "99.00", "999.0"))
buoy_1988 <- read_csv('https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1988.csv', na = c("99", "999", "99.00", "999.0"))
buoy_1989 <- read_csv('https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1989.csv', na = c("99", "999", "99.00", "999.0"))
buoy_1990 <- read_csv('https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1990.csv', na = c("99", "999", "99.00", "999.0"))




```

Answer: all the four codes read comma separated values (csv) files, the path/directory is the same, the character values to replace na are the same for all data files.
The codes differ in the file name extended by the year only (44013_"")


**2.2 Complete the skeleton of the for loop below, which uses the `str_c()` function to print out the path to the buoy 44013 data file from year `start` to `end`**


```{r,eval= F}

start <-1987

end <- 1992

for (year in start:end){
  
  path <- str_c("https://", "raw.githubusercontent.com/","nt246/", "NTRES6940-data-science/","master/", "datasets/", "buoydata/", "44013_", year, ".csv")
  
  print(path)
  
}


```





```{r}

##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1987.csv"
##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1988.csv"
##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1989.csv"
##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1990.csv"
##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1991.csv"
##[1] "https://raw.githubusercontent.com/nt246/NTRES6940-data-science/master/datasets/buoydata/44013_1992.csv"


```


**2.3 Complete the skeleton of the for loop below, which reads the buoy 4444013 data from year `start` to `end` and combine them together**


```{r}

start <-  1987

end <-  1992 

for (year in start:end){
  
  path <- str_c("https://", "raw.githubusercontent.com/","nt246/", "NTRES6940-data-science/","master/", "datasets/",
                "buoydata/", "44013_", year, ".csv")
  
  df<- read_csv(path)
  
  if(year == start){
    
    df_combined <- 
      
  } else {
    
    df_combined <- 
      
  }
}
```



**2.4 Building on the workflow that you used in 2.1 - 2.3, use a for loop to read in clean up, and summarize the buoy data from all years from 1987 to 1992 using a dplyr workflow.**


**Select** only columns `YY`(year), `MM` (month), `WVHT` (wave heights), `WTMP` (temperatures) and **rename** these columns to something understandable. **Summarize** monthly averaged wave heights and temperatures throughout the years in a tibble, and **plot the variation of these monthly averaged values through time** as shown below.

There are multiple ways to do this, and for this execercise, you may well combine all the raw data in a for loop and clean it up after the loop. In the next (**Optional**) excercise however, you will need to clean up the data in the loop before you can combine them. 

















