---
title: "Assignment 7: Iteration (for loops) and functions"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Instructions: please read through this before you begin**

* This homework is due by **10pm on Monday 11/09/20.**

* Please **reproduce this markdown template.** Pay attention to all the formatting in this file, including bullet points, bolded characters, inserted code chuncks, headings, text colors, blank lines, and etc.

* When a verbal response is needed, answer by editing the part in the R markdown template where it says "Write your response here".

* Have all your code embedded within the R markdown file, and show both your code and plots in the knitted markdown file.

* Use R Markdown functionalities to **hide messages and warnings when needed**. (Suggestion:messages and warnings can often be informative and important, so please examine them carefully and only turn them off when you finish the exercise).

* Please name your R markdown file `assignment_7.Rmd` and the knitted markdown file `assignment_7.md`. Please upload both files using your personal GitHub repository for this class.


**Acknowledgments:** Exercise 1 has been adapted (with permission) from the **[Data Carpentry for Biologists semester course](https://datacarpentry.org/semester-biology/exercises/Functions-use-and-modify-Python/)** and exercises 2 and 3 are adapted (with permission) from lain Carmichael's course **[STOR 390:Introduction Data Science](https://idc9.github.io/stor390/#course_material).**

To start, first load all the required packages with the following code. Install them if they are not installed yet.


```{r, message = FALSE}

library(tidyverse)

```


### **Exercise 1: Body mass estimation using vectorization vs. for loop**

There are two major types of approaches to perform multiple operations in R:**[vectorization](https://swcarpentry.github.io/r-novice-gapminder/09-vectorization/)** and for loops. As a simple to calculate the sum of two vectors, `x` and `y`, the syntax for vectorization is simply `z <- x + y`. With this, the computer will be able to perform the same operation to each element of x and y vector **simultaneously.**

The for loop approach to achieve the same task (create a new vector z that is the sum of vectors x and y), on the other hand, takes the following form:

```{r, eval=FALSE}

z[i] <- NULL

for (i in 1:length(x)){
  
  z[i] <- x[i] +y[i]
  
  }

```

In this case, the computer loops through each element of x and y and performs the operation **sequentially**, resulting in a significantly longer runtime. Let's now try to quantify this difference in runtime in this exercise.




**1.1**










